---
title: "GIS Workshop Pt1"
author: "M.Cortese"
date: "2025-07-10"
output: html_document
Notes: Adopted from the terra vingette. See for more tutorials https://rspatial.org/pkg/index.html
---

Learning Goals:
Projections & coordinate systems
Loading spatial data
Building new layers 
Editing data (and metadata)
Exporting data
Making and exporting basic maps

This will cover processes in terra to avoide conflicts between multiple package format. If you start using other common geospatial packages like 'raster' you may have to start switching between file formats 'spatRaster' to 'rasterLayer'

```{r setup}
#load packages
library(terra)
library(ggspatial)
library(tidyterra)
library(ggplot2)
```

First we want to load in some spatial data that already exists. This could include point data, lines, polygons, or raster data. 

```{r loading spatial data}
#points
fish <- read.csv("data/Occurrence.csv")
lat <- fish$decimallatitude
long <- fish$decimallongitude
lonlat <- cbind(long, lat)
snapper <- vect(lonlat)
plot(snapper)

#polygon
fish_range <- vect("data/Species_Range.shp")
plot(fish_range)

#raster
pH <- rast("data/Mean_PH.tif")
plot(pH)

#raster from netCDF
temp <- rast('data/ltMax_temp.nc')
plot(temp)

```

Or we may want to just make our own spatial data. This is useful if you are entering points by hand,need a layer to add other data into later, or to test a method

```{r Building new data}
crs <- "+proj=longlat +datum=WGS84"

#point data
long <- c(-76.5, -80.2, -75.2, -77.4, -79.1, -74.0, -73.7, -76.9, -70.3, -71.6)
lat <- c(38.8, 42.6, 38.9, 42.1, 36.7, 38.1, 36.2, 39, 41.6, 37.9)
lonlat <- cbind(long, lat)
point <- vect(lonlat, crs= crs) #make sure to include the crs since it wont autogenerate one like in ArcMap
plot(point)

#Polygon
long <- c( -75,  -76, -75.5, -73.5,  -74.5)
lat <- c(38.9,  38.3, 38.2, 38.1,  38.7)
lonlat <- cbind(id=1, part=1, long, lat)
polygon <- vect(lonlat, type="polygons", crs=crs) # can swap type= "lines" for a line rahter than a polygon
plot(polygon)

#Raster
r <- rast(ncol=10, nrow=10, xmin=-150, xmax=-50, ymin=50, ymax=100, crs = crs) # this builds an empty raster to then put other data into
values(r) <- 1:ncell(r) #assigns sequential cell numbers as the value
plot(r)
```

Then we may want to edit that data. Below we will subset our point data to the polygon, and crop our rasters two different ways

```{r Editing data}
#clip rasters
pH_Crop <- crop(pH, fish_range, mask = TRUE) #first crop one raster to the polygon extent using the mask function
plot(pH_Crop)

temp_Crop <- crop(temp, pH_Crop) #notice this doesn't do the mask with exact outline, it just crops to the outer boundary
plot(temp_Crop)

occurance_crop <- crop(snapper, fish_range)
plot(occurance_crop)

```

Don't forget you may need to change a coordinate system
```{r changing coordinate systems}
#project to a new coordinate system
newcrs <- "+proj=robin +datum=WGS84"
transformed <- terra::project(r, newcrs)
plot(transformed)

#assign coordinate system for the first time
crs(temp_Crop) <- crs
crs(occurance_crop) <- crs
```

Next we want to make a nice map with all of our data!
```{r mapping}
#naming to export later
snapper_map <- ggplot() +
  geom_spatraster(data = temp_Crop)+ #add raster
  geom_spatvector(data = occurance_crop, col = "black") + #add point data
  annotation_north_arrow(location = "tr", which_north = "true", pad_x = unit(0.1, "in"), pad_y = unit(0.5, "in"), style = north_arrow_minimal) + #add north arrow
  annotation_scale(location = "bl", width_hint = 0.5,pad_x = unit(0.2, "in"), pad_y = unit(0.3, "in")) + #add scale bar
 labs(fill = "Longterm Max SST",
      title = "Gray Snapper" ) 

snapper_map

```

Last we need to save our map and data

```{r saving data}
#save our cropped snapper data
writeVector(occurance_crop, "data/cropped_snapper.shp", overwrite=TRUE)
#save our cropped temp data
writeRaster(temp_Crop, "data/cropped_temp.tif", overwrite=TRUE)
#save map
ggsave("snapper_map.png", plot = snapper_map, path = "data/")

```

